GEE models

### 2017-10-16 doc/submission/reviewer-response-2.md

    - *"it is a pity that the analysis of fatty acids composition in
    phospholipids and serum cholesteryl esters have been published in a separate
    paper, altogether this would add very much to the total interpretation."*

        While we agree with the reviewer that inclusion of the phospholipid and
        cholesteryl ester (as well as the triacylglycerol) fractions would have
        added to the interpretation, the volume and complexity of the analyses
        would not have been publishable given current journal word and figure
        count limits. This was the reason we decided to publish the fractions in
        separate manuscripts, to maintain a clear and focused narrative within
        the constraints of the journals limits.

    - *"some discussion on what determines plasma fatty acid profile may be
    useful, beside spillover also (selective) release by endogenous lipolysis.
    To what extent does it reflect dietary composition, was habitual fat intake
    taken into account?"*

        Fasting NEFA derives from adipose tissue, which contains fat stores from
        both diet and de novo lipogenesis. Besides potential dysfunction in the
        adipose tissue for the release of NEFA, other factors that could determine
        the NEFA profile include genetics and long term dietary fat intake
        (adipose tissue fat stores have a half-life of ~1.5 years). These other
        factors were not considered and discussed in the present paper due to
        the difficulty in measuring long term diet accurately enough on the
        specific type of fat consumed, the lack of genetic information contained
        in the cohort, and the fact that disentangling the specific contribution
        of de novo lipogenesis from dietary fat on adipose tissue stores is well
        outside the scope of this paper.

### 2017-10-17 doc/submission/reviewer-response-2.md

    3. *"Please add the number of observations for each clinic visit in Table 1."*

        As requested, this has been added lto Table 1.

    4. *"Please add the number of incident and prevalent diabetes cases to the
    respective exclusions in ESM Figure 1. For example: n=736 had baseline data
    collected, n=XX diabetes cases were excluded; n=619 had 3-year data, n=XX
    incident diabetes cases were excluded..."*

        The exact number of excluded cases of diabetes is dependent on this analysis
        and doesn't reflect the main PROMISE cohort. Since the CONSORT diagram 
        represents the main cohort, we included the number of excluded diabetes
        cases in the text of the Results section since the text reflects the 
        present analysis.

### 2017-10-17 doc/manuscript.Rmd
started text on excluded cases at each visit
added other refs on beta cell decline

For this study, there were `r female$npct` females and `r european$npct` had
European-ancestry (see `r cite_t("basic")`). The mean follow-up time was 
`r calc_followup_time()` years, where `r calc_pct_full_visits()`% of 
participants attended all three visits.
The average age of the participants at baseline was `r aide::average(ds0yr$Age)`
years (`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` kg/m^2^ (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r famhist$npct`, had a family history of diabetes. Insulin
sensitivity and beta cell function measures showed a significant median
decline from baseline to 6-years of between `r chg_outcomes$chg` 
(p`r chg_outcomes$p`, based on a time-adjusted GEE model) in this analysis.
Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
and `r calc_incid_dysgly("ConvertPreDM")`incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
cases were excluded from GEE analyses (follow-up data were included until date
of diagnosis). Certain medications (e.g. lipid medications such as statins or
fibrates) could influence the association between NEFA and the outcomes.
There were `r meds_use[1]` and `r meds_use[2]` participants who used lipid and
cholesterol modifying medications at the baseline and 6-year visit,
respectively. Use of these medications did not influence NEFA composition, the
outcomes, or the GEE models (data not shown).

@@

The role of total NEFA in the etiology of diabetes is well-documented.
Epidemiological studies have shown that higher NEFA associate with lower insulin
secretion and a higher risk for developing diabetes 
[@Paolisso1995a;@Salgin2012a;@Pankow2004a;@Charles1997a]. In a cross-sectional analysis of the
RISCK cohort, total NEFA had a negative association with insulin sensitivity and
a particularly strong negative association with beta cell function
[@Johns2014a]. Experimentally, several potential mechanisms have been elucidated
for the role of NEFA on beta cell function, particularly for palmitic acid.
Prolonged exposure to elevated NEFA can induce apoptosis in the beta cells,
possibly through endoplasmic reticulum stress, formation of ceramides, and
generation of nitric oxide, as well as impairment of proinsulin production and
mitochondrial function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a]. The
present analysis is the first study to our knowledge to examine the longitudinal
association of a broad spectrum of individual NEFA on beta cell function in a
large cohort. We found that there was a strong signal of higher total NEFA,
palmitic acid, oleic acid, and linoleic acid modeled as concentrations with
lower beta cell function. No fatty acid as a mol% had associations with
beta cell function. Taken together, these results suggest that it is the
absolute size of the circulating NEFA fraction, irrespective of any specific 
composition of fatty acids, that is responsible for the hypothesized lipotoxic 
effects of NEFA on the beta cells.

>>>>

For this study, there were `r female$npct` females and `r european$npct` had
European-ancestry (see `r cite_t("basic")`). The mean follow-up time was 
`r calc_followup_time()` years, where `r calc_pct_full_visits()`% of 
participants attended all three visits.
The average age of the participants at baseline was `r aide::average(ds0yr$Age)`
years (`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` kg/m^2^ (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r famhist$npct`, had a family history of diabetes. Insulin
sensitivity and beta cell function measures showed a significant median
decline from baseline to 6-years of between `r chg_outcomes$chg` 
(p`r chg_outcomes$p`, based on a time-adjusted GEE model) in this analysis.
Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
and `r calc_incid_dysgly("ConvertPreDM")`incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
cases were excluded from GEE analyses (n={{num}} at 3rd year, n=
{{num}} at 6th year). Follow-up data were included until date of
diagnosis.
Certain medications (e.g. lipid medications such as statins or
fibrates) could influence the association between NEFA and the outcomes.
There were `r meds_use[1]` and `r meds_use[2]` participants who used lipid and
cholesterol modifying medications at the baseline and 6-year visit,
respectively. Use of these medications did not influence NEFA composition, the
outcomes, or the GEE models (data not shown).

@@

The role of total NEFA in the etiology of diabetes is well-documented.
Epidemiological studies have shown that higher NEFA associate with lower insulin
secretion and a higher risk for developing diabetes 
[@Paolisso1995a;@Salgin2012a;@Pankow2004a;@Charles1997a]. In a cross-sectional analysis of the
RISCK cohort, total NEFA had a negative association with insulin sensitivity and
a particularly strong negative association with beta cell function
[@Johns2014a]. Another cross-sectional study found a null association of
total NEFA on beta cell function [@Rebelos2015a], however they used C-peptide to
estimate beta cell function.
Experimentally, several potential mechanisms have been elucidated
for the role of NEFA on beta cell function, particularly for palmitic acid.
Prolonged exposure to elevated NEFA can induce apoptosis in the beta cells,
possibly through endoplasmic reticulum stress, formation of ceramides, and
generation of nitric oxide, as well as impairment of proinsulin production and
mitochondrial function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a]. 
One longitudinal studies also found an inverse association between total NEFA
and beta cell function over 10 years in 77 Japanese patients with type 2
diabetes [@Morita2012a], though they used C-peptide to estimate beta cell
function. The
present analysis is the first study to our knowledge to examine the longitudinal
association of a broad spectrum of individual NEFA on beta cell function in a
large cohort. We found that there was a strong signal of higher total NEFA,
palmitic acid, oleic acid, and linoleic acid modeled as concentrations with
lower beta cell function. No fatty acid as a mol% had associations with
beta cell function. Taken together, these results suggest that it is the
absolute size of the circulating NEFA fraction, irrespective of any specific 
composition of fatty acids, that is responsible for the hypothesized lipotoxic 
effects of NEFA on the beta cells.
